Income Inequality is revealed in Torontonian Apartment Maintenance

### 2022-02-07 outputs/paper/paper.Rmd
Last changes on some grammatical errors.

We obtained the data set from the City of Toronto Open Data Portal [@citewebsite] with `opendatatoronto` [@citeopentoronto], and cleaned and analyzed the data set using `R` [@citeR], `dplyr`, [@citedplyr] and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit] and `kableExtra` [@citekable]. We first created tables and graphs that showed the data. We then went into showing the relevant information for apartment types and wards using box plots, linear model graphs and percentage bar charts. These all highlighted inequality in housing evaluation scores.

@@

The data for this report was gotten from the Apartment Building Evaluation [@citedata] of the Toronto Open Data Portal [@citewebsite] using the R package opendatatoronto [@citeopentoronto]. This data is compiled by RentSafeTO and was last accessed on January 22nd, 2022. Rent Safe TO collects data on apartments in the Greater Toronto Area (GTA) that are 3 or more stories tall or had 10 more units [@torontoeval], and response is mandatory for qualifying apartments, i.e the population. RentSafeTO collects the data by setting up an appointment with the landlord of the apartment block for inspection, and take notes and photographs of their inspection to upload to the City’s mobile investigation application. Later, the total building evaluation score is calculated and made available to review and post on the Tenant Notification Board [@torontoeval]. The raw data includes 9758 apartment blocks from the GTA and their categorical characteristics such as location and Type of property, as well as numeric characteristics such as Security,amenities and state of common areas. It also includes an overall score over 100 for each apartment block, and there are 40 variables in total.

@@

We included a variable in our data frame if there was anything important that made it stand out for our analysis. We omitted most variables in the original set as their direct effects were not important to the analysis, and the total score variable encompassed all of their individual scores. we first removed _id and RSN as they were merely ID variables. We also excluded the numeric ward variable as the ward names variable, which contains the 25 Toronto wards, was easier to use for identification. We included year built as we wanted to see the changes in apartment evaluation between older and newer apartments. We included score as it was the main determinant of maintenance standards. We also included Security (which measures the condition of security systems of the apartment) to observe the the direct relationship and differences in security between different groups of apartments. We then included property type, which states whether an apartment was privately owned, owned by the TCHC, or owned by some other non-private institution, as our analysis focuses on social/affordable housing. The final data frame included 493 observations with 5 variables.

@@

We are interested in the relationships between apartment Type, Ward,Score and Security, as well as the trend of Score over time. These relationships show a measure of income inequality in housing, and have possible implications for public policy. The data, (Table \@ref(tab:tabone)), (Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `kableextra` [@citekable], `knitr` [@citeknit] and `ggplot2` [@citeggplot] respectively are shown below.

@@

(Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) show the spread and distribution of apartments scores and their security levels respectively. We see that the data for Scores seems to be unimodal and normally distributed. There is an outlier of a very low apartment score, but it is not enough to skew the data to the left.The median apartment score seems to be the same or a slightly above the mean apartment score, as there seems to be more apartments on the lower end than there are on the higher end. This indicates that there are many below average apartments. We can see from the figures that the average building has good maintenance and a high standard of security. However as there are still quite a lot of lower security apartments, it seems that the mean level of security may be lower than the median. Policy is needed to improve security for these apartments, and the observation that some evaluation Scores are quite lower than most are more thoroughly scrutinized later in the report.

@@

```{r figsix,out.width="90%", fig.cap = "Percentage of apartments in each ward with scores in a certain range", echo = FALSE }
rent_clean %>%
  mutate(Score = 
           case_when(
             Score <= 50 ~ "Less than 50",
             Score <= 60 ~ "50 to 60",
             Score <= 70 ~ "60 to 70",
             Score <= 80 ~ "70 to 80",
             Score <= 100 ~ "80 or more",
             TRUE ~ "None"
             ) # group scores into separate categories
  ) %>%
  ggplot(data=.)+
  theme_minimal()+
  coord_flip()+
  geom_bar(mapping = aes(x = Ward, fill = Score),position="fill")+ # coloring the different categories
  scale_y_continuous(labels = scales::percent)+ # getting the percentage of scores within each ward
  scale_color_manual("YlOrRd")+
  labs(x = "Ward",
       y = " Percentage of Apartments in ward",
       color = "Evaluation Score") 
```

@@

The report has shown a disparity in priorities for apartment maintenance and upkeep between public and private apartments, as well as general housing inequality as those in social housing owned by the TCHC do not have the high apartment quality or security they deserve [@rights]. Regarding these apartments that have less security, it is paradoxical as these people often live in more crime ridden areas that require the extra security [@Stretesky2004]. This brings into question if their own lives are considered, as well as the failings of public landlords and the TCHC to provide security akin to those of private apartments. Lower income people often do not have money to protect themselves like the wealthy, so they need to rely on publicly provided assistance.

The data showed that Scarborough North, which has a 90% visible minority population with an average yearly income of 78,892 dollars as of 2016 compared to the Toronto average of 102,721 dollars [@torontoward], only has highly scored apartment buildings. If this is not due to a sampling error, it is hopeful that apartment scores for other wards with similar incomes increase in the future. However low income residents have concerns that a high level of apartment security, or highly rated apartments in an otherwise lower income area can make the area more attractive to wealthier people. This could lead to gentrification, which pushes the original low income inhabitants out, which is a problem with the already tenuous social housing situation for these marginalized groups [@August2014] and [@Crosby2020].

@@

There are some data collection problems with this data set. RentSafeTO collects data only for apartment blocks with 3 or more stories or ten or more units[@torontoeval], meaning that some buildings that are in dire need of maintenance may be left behind as the organization simply does not collect data for them, even though most people may see them as apartments (i.e a three story block of 9 flats). This leaves out a big piece of the full picture on apartment quality, and may also be why some known wealthier wards are getting higher average scores as the low income housing there does not count. It also leaves out the full effect of income inequality as lower income people are now more likely to live in such types of "apartments" than the wealthy [@Musterd2014]. For apartments that do qualify for data collection, the grading is a scheduled event organized with the landlord ahead of time [@torontoeval]. This means that scores presented may not be an entirely accurate measure of the general upkeep of a building as landlords who neglect their property can hurriedly make repairs and straighten out the apartment block before the inspector/enforcement officer arrives. It may thus be necessary for inspectors to come up semi-unannounced to get the true evaluation of the building.

@@

Private housing is now in itself getting harder to afford due to rising apartment costs, forcing many so called middle class people into social housing or lower rated private apartments. The effect of rising prices and the movement of people that would not have been considered low income into apartments in lower income wards or social housing apartments, causes a causing a higher demand for these apartments and thus causing a larger strain on already heavily strained available social housing funds [@irwin].

@@

For next steps we could use the geo-location variable that was omitted for this report to see if the scores for apartments on boundaries between poorer or richer wards had the negative bias of the poor wards or positive bias of the rich. We could also use this variable for a spatial comparison in contrast to the graphical comparison used in the report. We could also include apartment blocks that did not have all amenities in the comparison of income and scores to further observe evidence of inequality, as apartments in poor areas tend to have complete absence of some amenities.

>>>>

We obtained the data set from the City of Toronto Open Data Portal [@citewebsite] with `opendatatoronto` [@citeopentoronto], and cleaned and analyzed the data set using `R` [@citeR], `dplyr`, [@citedplyr] and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr] and `kableExtra` [@citekable]. We first created tables and graphs that showed the data. We then went into showing the relevant information for apartment types and wards using box plots, linear model graphs and percentage bar charts. These all highlighted inequality in housing evaluation scores.

@@

The data for this report was gotten from the Apartment Building Evaluation [@citedata] of the Toronto Open Data Portal [@citewebsite] using the R package `opendatatoronto` [@citeopentoronto]. This data is compiled by RentSafeTO and was last accessed on January 22nd, 2022. Rent Safe TO collects data on apartments in the Greater Toronto Area (GTA) that are 3 or more stories tall or had 10 more units [@torontoeval], and response is mandatory for qualifying apartments, i.e the population. RentSafeTO collects the data by setting up an appointment with the landlord of the apartment block for inspection, takes notes and photographs of their inspection to upload to the City’s mobile investigation application. Later, the total building evaluation score is calculated and made available to review and post on the Tenant Notification Board [@torontoeval]. The raw data includes 9758 apartment blocks from the GTA and their categorical characteristics such as location and Type of property, as well as numeric characteristics such as Security,amenities and state of common areas. It also includes an overall score over 100 for each apartment block, and there are 40 variables in total.

@@

We included a variable in our data frame if there was anything important that made it stand out for our analysis. We omitted most variables in the original set as their direct effects were not important to the analysis, and the total score variable encompassed all of their individual scores. We first removed _id and RSN as they were merely ID variables. We also excluded the numeric ward variable as the ward names variable, which contains the 25 Toronto wards, was easier to use for identification. We included year built as we wanted to see the changes in apartment evaluation between older and newer apartments. We included score as it was the main determinant of maintenance standards. We also included Security (which measures the condition of security systems of the apartment) to observe the the direct relationship and differences in security between different groups of apartments. We then included property type, which states whether an apartment was privately owned, owned by the TCHC, or owned by some other non-private institution, as our analysis focuses on social/affordable housing. The final data frame included 493 observations with 5 variables.

@@

We are interested in the relationships between apartment Type, Ward,Score and Security, as well as the trend of Score over time. These relationships show a measure of income inequality in housing, and have possible implications for public policy. A glimpse of the data, (Table \@ref(tab:tabone)), using `kableextra` [@citekable], and `knitr` [@citeknit] are shown below.

@@

(Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of apartments scores and their security levels respectively. We see that the data for Scores seems to be unimodal and normally distributed. There is an outlier of a very low apartment score, but it is not enough to skew the data to the left. The median apartment score seems to be the same or a bit above the mean apartment score, as there seems to be more apartments on the lower end than there are on the higher end. This indicates that there are many below average apartments. We can see from the figures that the average building has good maintenance and a high standard of security. However as there are still quite a lot of lower security apartments, it seems that the mean level of security may be lower than the median. Policy is needed to improve security for these apartments, and the observation that some evaluation Scores are quite lower than most are more thoroughly scrutinized later in the report.

@@

```{r figsix,out.width="90%", fig.cap = "Percentage of apartments in each ward with scores in a certain range", echo = FALSE }
rent_clean %>%
  mutate(Score = 
           case_when(
             Score <= 50 ~ "Less than 50",
             Score <= 60 ~ "50 to 60",
             Score <= 70 ~ "60 to 70",
             Score <= 80 ~ "70 to 80",
             Score <= 100 ~ "80 or more",
             TRUE ~ "None"
             ) # group scores into separate categories
  ) %>%
  ggplot(data=.)+
  theme_minimal()+
  coord_flip()+
  geom_bar(mapping = aes(x = Ward, fill = Score),position="fill")+ # coloring the different categories
  scale_y_continuous(labels = scales::percent)+ # getting the percentage of scores within each ward
  guides(fill=guide_legend(title = "Score Group"))+
  labs(x = "Ward",
       y = " Percentage of Apartments in ward",
       color = "Evaluation Score") 
```

@@

The report has shown a disparity in priorities in apartment maintenance and upkeep between public and private apartments, as well as general housing inequality as those in social housing owned by the TCHC do not have the high apartment quality or security they deserve [@rights]. Regarding these apartments that have less security, it is paradoxical as these people often live in more crime ridden areas that require the extra security [@Stretesky2004]. Lower income people often do not have money to protect themselves like the wealthy, so they need to rely on publicly provided assistance. This brings into question if their own lives are considered, as well as the failings of public landlords and the TCHC to provide security akin to those of private apartments.

The data showed that Scarborough North, which has a 90% visible minority population with an average yearly income of 78,892 dollars as of 2016 compared to the Toronto average of 102,721 dollars [@torontoward], only has highly scored apartment buildings. If this is not due to a sampling problem, it is hopeful that apartment scores for other wards with similar incomes increase in the future. However low income residents have concerns that a high level of apartment security, or highly rated apartments in an otherwise lower income area can make the area more attractive to wealthier people. This could lead to gentrification, which pushes the original low income inhabitants out, which is a problem with the already tenuous social housing situation for these marginalized groups [@August2014] and [@Crosby2020].

@@

There are some data collection problems with this data set. RentSafeTO collects data only for apartment blocks with 3 or more stories or ten or more units [@torontoeval], meaning that some buildings that are in dire need of maintenance may be left behind as the organization simply does not collect data for them, even though most people may see them as apartments (i.e a three story block of 9 flats). This leaves out a big piece of the full picture on apartment quality, and may also be why some known wealthier wards are getting higher average scores as the low income housing there does not count. It also leaves out the full effect of income inequality as lower income people are now more likely to live in such types of "apartments" than the wealthy [@Musterd2014]. For apartments that do qualify for data collection, the grading is a scheduled event organized with the landlord ahead of time [@torontoeval]. This means that scores presented may not be an entirely accurate measure of the general upkeep of a building as landlords who neglect their property can hurriedly make repairs and straighten out the apartment block before the inspector/enforcement officer arrives. It may thus be necessary for inspectors to come up semi-unannounced to get the true evaluation of the building.

@@

Private housing is now itself getting harder to afford due to rising apartment costs, forcing many so called middle class people into social housing or lower rated private apartments. The effect of rising prices has led to the movement of people that would not have been considered low income into social housing apartments or apartments in lower income wards, causing a higher demand for these apartments and thus leading to a larger strain on already heavily strained available social housing funds [@irwin].

@@

For next steps we could use the geo-location variable that was omitted for this report to see if the scores for apartments on boundaries between poorer or richer wards had the negative bias of the poor wards or positive bias of the rich. We could also use this variable for a spatial comparison in contrast to the graphical comparison used in the report. We could then include apartment blocks that did not have all amenities in the comparison of income and scores to further observe evidence of inequality, as apartments in poor areas tend to have complete absence of some amenities.
